---
title: "Lab 1: Displaying correlated data"
subtitle: "EPIB 627, Fall 2020"
author: "Iris Ganser"
date: "09/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Considering the design of the study, how does the clustering arise?
The clustering in the study comes from two factors:  
1. Data were collected at many hospitals. Babies treated at the same hospital might be more similar to each other than babies treated at different hospitals.   
2. Data were collected longitudinally, i.e. at many different timepoints on the same babies. 

## 2. Are subjects within the same hospital more alike than others? 
(a) Make a plot to visualize the hospital-level variability at time=0 and time=12. 
```{r}
# load data
probit <- read.csv("probit.csv")
probit$HOSP <- as.factor(probit$HOSP)
probit$SUBJ <- as.factor(probit$SUBJ)

# load packages
library(ggplot2)
library(dplyr)
library(nlme)

# hospital variability at time = 0
ggplot(filter(probit, TIME == 0), aes(x = HOSP, y = WAZ)) +
  geom_boxplot() +
  labs(x = "Hospital", y = "Weight for Age Z-score (WAZ)", 
       title = "Hospital differences in WAZ at time = 0")

# hospital variability at time = 12
ggplot(filter(probit, TIME == 12), aes(x = HOSP, y = WAZ)) +
  geom_boxplot() +
  labs(x = "Hospital", y = "Weight for Age Z-score (WAZ)", 
       title = "Hospital differences in WAZ at time = 12")
```

(b) Calculate the ICC for the hospital level clustering at birth (time=0) and at 12 months
(time=12).
```{r}
# model with a random intercept per hospital to account for hospital clustering and further compute ICC:
lme0 <- lme(WAZ ~ 1, random = ~1|HOSP, data = filter(probit, TIME == 0))
lme12 <- lme(WAZ ~ 1, random = ~1|HOSP, data = filter(probit, TIME == 12))

# VarCorr function extracts the components we are interested in
icc0 <- as.numeric(VarCorr(lme0)[1,1])/
  (as.numeric(VarCorr(lme0)[1,1]) + as.numeric(VarCorr(lme0)[2,1]))
icc0

icc12 <- as.numeric(VarCorr(lme12)[1,1])/
  (as.numeric(VarCorr(lme12)[1,1]) + as.numeric(VarCorr(lme12)[2,1]))
icc12
```

Almost no hospital clustering is observable at time = 0 (ICC = 0.008). At time = 12, hospital clustering is a bit higher (ICC = 0.015).

## 3. Are observations on the same baby more alike than others?
(a) Make a plot to visualize the baby-level variability (for hospital 1 babies)
```{r}
ggplot(data = filter(probit, HOSP == 1), aes(x = SUBJ, y = WAZ)) +
  geom_boxplot() +
  labs(x = "Baby ID", y = "Weight for Age Z-score (WAZ)", 
       title = "Boxplots per baby in hospital #1")
```

Visually, there are a lot of differences between babies. Makes sense!

(b) Calculate the ICC for the baby-level clustering for babies from hospital 1.
```{r}
lme_baby <- lme(WAZ ~ 1, random = ~1|SUBJ, 
                data = filter(probit, HOSP == 1))
icc_baby <- as.numeric(VarCorr(lme_baby)[1,1])/
  (as.numeric(VarCorr(lme_baby)[1,1]) + as.numeric(VarCorr(lme_baby)[2,1]))
icc_baby
```
There is a high intra-class correlation (0.62), meaning that 62% of the variability
between observations is due to the babies. 

## 4. Which is bigger, the ICC for babies or for hospitals? What happens to the hospital ICC
between time=0 and time=12? How do the boxplots for babies compare to the boxplot for
hospitals? Interpret the ICCs.

The ICC for babies is much higher, meaning there are more individual differences. ICC for hospitals between time=0 and time=12 increases, meaning differences between hospitals increase over the first year of life of the babies. Compared to the boxplots for hospitals, the boxplots for babies show much more heterogeneity. 0.8% of the variability between observations is due to hospitals whereas 62% of the variability between observations is due to the babies. 

## 5. Explore the baby-level variability using a trajectory plot, a cross correlation plot, variogram, and correlation matrix for one hospital!
```{r}
# filter data for hospital 1
hosp1 <- filter(probit, HOSP ==1)
hosp1$SUBJ <- factor(hosp1$SUBJ)
baby <- as.numeric(sample(levels(hosp1$SUBJ), size = 25, replace = FALSE))
hosp1 <- filter(hosp1, SUBJ %in% baby)

# trajectory plot for 25 random babies
ggplot(data = hosp1, aes(x = TIME, y = WAZ, group = SUBJ)) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 12, 1)) +
  labs(title = "Trajectories for 25 random babies in hospital #1")

# cross correlation plot
# regress outcome on predictors and extract residuals
hosp1 <- filter(probit, HOSP ==1)
WAZlm <- lm(WAZ ~ WAZ0 + TIME, 
            data = hosp1)
resWAZ <- data.frame(SUBJ = hosp1$SUBJ, res = residuals(WAZlm))
resWAZ <- resWAZ %>% group_by(SUBJ) %>% mutate(observation = row_number())

# create wide dataset
library(tidyr)
resWAZ_wide <- pivot_wider(resWAZ, names_from = observation, values_from = res)

# cross correlation plot
chart.Correlation(resWAZ_wide[, 2:8], histogram=TRUE, na.action = na.omit)


# variogram
# replace subject numbers with continuous numbers from 1 to 94
resWAZ_wide$id <- 1:94

hosp1 <- left_join(hosp1, resWAZ_wide, by = "SUBJ") %>% 
  select(-c(36:42))
# fit linear model with only time as predictor
WAZlmT <- lm(WAZ ~ TIME, data = hosp1)
#calculate variance values
myval <- outer(resid(WAZlmT), resid(WAZlmT), function(x, y) ((x - y)^2)/2)

# need the number of observations per subject as we will replace diagonal values by NA in each submatrix from an individual
obs.byid<-table(hosp1$id)

i<-1
j<-1

# change diagonal values (the zeros) by NA in submatrices, as these were not calculated among two different individuals
while(i <= dim(myval)[1]){
  myval[i:(i+(obs.byid[j]-1)),i:(i+(obs.byid[j]-1))]<-NA
  i<-i+obs.byid[j]
  j<-j+1
}

#total variation
process.var <- mean(myval, na.rm=TRUE)

# linear model with explanatory variable
fm3 <- gls(WAZ ~ TIME, data=hosp1)

# run the variogram function, the form tells R about how the data are clustered
Var1<-Variogram(fm3, form = ~ TIME |id, collapse="none", resType="response")

# now a scatter plot
plot(Var1[,2], Var1[,1],ylim=c(0,2*process.var),xlab="Time Lag",ylab="Variogram")

# plsmo function plots smoothed estimates
plsmo(Var1[,2], Var1[,1],add=TRUE,lty=2,"raw",col="red")

# add a horizontal line for the process variance
abline(h=process.var)

# or if you prefer, a loess curve
Var2<-Var1[order(Var1[,2],Var1[,1]),]
lines(wtd.loess.noiter(Var2[,2], Var2[,1]),col="blue",lwd=2)
title(main="variogram -- horizontal line=process variance", sub="red dashed line= unsmoothed variogram, blue line =smoothed variogram")

## now include WAZ0 in model
WAZlmT <- lm(WAZ ~ TIME + WAZ0, data = hosp1)
#calculate variance values
myval <- outer(resid(WAZlmT), resid(WAZlmT), function(x, y) ((x - y)^2)/2)

obs.byid<-table(hosp1$id)

i<-1
j<-1

# change diagonal values (the zeros) by NA in submatrices, as these were not calculated among two different individuals
while(i <= dim(myval)[1]){
  myval[i:(i+(obs.byid[j]-1)),i:(i+(obs.byid[j]-1))]<-NA
  i<-i+obs.byid[j]
  j<-j+1
}

#total variation
process.var <- mean(myval, na.rm=TRUE)

# linear model with explanatory variable
fm3 <- gls(WAZ ~ TIME + WAZ0, data=hosp1)

# run the variogram function, the form tells R about how the data are clustered
Var1<-Variogram(fm3, form = ~ TIME |id, collapse="none", resType="response")

# now a scatter plot
plot(Var1[,2], Var1[,1],ylim=c(0,5*process.var),xlab="Time Lag",ylab="Variogram")

# plsmo function plots smoothed estimates
plsmo(Var1[,2], Var1[,1],add=TRUE,lty=2,"raw",col="red")

# add a horizontal line for the process variance
abline(h=process.var)

Var2<-Var1[order(Var1[,2],Var1[,1]),]
lines(wtd.loess.noiter(Var2[,2], Var2[,1]),col="blue",lwd=2)
title(main="variogram -- horizontal line=process variance", sub="red dashed line= unsmoothed variogram, blue line =smoothed variogram")
```

