---
title: 'Lab 2: Mixed models for continuous outcomes'
subtitle: "EPIB 627, Fall 2020"
author: "Iris Ganser"
date: "17/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fit the 4 models performed in class, i.e. i) a simple linear regression model, ii) a random
intercept model, iii) a random slope model with independent random coefficients, iv) a random
slope model with correlated random coefficients.
```{r}
library(lme4)
exam <- read.csv("~/PhD/Courses/EPIB627/Labs/627labs/exam.csv")
str(exam)

# simple linear regression model
lm <- lm(normexam ~ standLRT, data = exam)
summary(lm)

# random intercept model
ri <- lmer(normexam ~ standLRT + (1|school), data = exam)
summary(ri)

# random slope model
rs <- lmer(normexam ~ standLRT + (1|school) + (-1 + standLRT|school), data = exam)
summary(rs)

# random slope model with correlated random coefficients
rs_cor <- lmer(normexam ~ standLRT + (1 + standLRT|school), data = exam)
summary(rs_cor)
```


## Compare the results for the four models with respect to: i) fixed effect estimates, ii) fixed effect standard errors, iii) AIC, iv) Correlation between 2 observations, v) Number of parameters to be estimated
```{r}
# fixed effect estimates
coef(lm)
fixef(ri)
fixef(rs)
fixef(rs_cor)

# fixed effects standard errors

# AIC
AIC(lm)
AIC(ri)
AIC(rs)
AIC(rs_cor)

# correlation between two observations
Vcov <- vcov(ri, useScale = FALSE)

# number of parameters to be estimated
```

## Try to reproduce the figures from the lecture: i) plots of random effects, ii) estimated regression lines, iii) graphs to check the model's assumptions
```{r}
# plots of random effects
coef_ri <- coef(ri)$school
plot(x = 0, y = 0)
for(i in 1:65){
  abline(coef_ri[i, 1], coef_ri[i, 2], col = "darkblue")
}
abline(h = 0,  lty = 2)
abline(v = 0, lty = 2)
```

