---
title: "lab4"
subtitle: "EPIB 627, Fall 2020"
author: "Iris Ganser"
date: "01/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Create variables for the following representations of time:  
(a) CD4 declines linearly with time   
(b) CD4 varies linearly, drops at seroconversion then declines linearly (i.e. there is a " jump" at seroconversion) 
i. Same slope before and after (CD4 declines)  
ii. Different slopes (increases first, and then decreases)  
(c) CD4 declines linearly until seroconversion, then declines more steeply after (i.e. a "bend"
at seroconversion, but no discontinuity)  
```{r}
### test for variable coding
time <- seq(-5, 5, 0.1)

ya <- 5 - 4*time
ybi <- 5 - 4*time - 3*I(time >= 0)
ybii <- 5 + 1.5*time - 2*time*I(time >= 0) - 3*I(time >= 0)
yc <- 5 - 4*time - 3*time*I(time >= 0)

plot(time, ya, type = "l", lwd = 3)
lines(time, ybi, col = 2, lwd = 3)
lines(time, ybii, col = 3, lwd = 3)
lines(time, yc, col = 4, lwd = 3)
```


```{r}
cd4 <- read.csv("cd4.csv")
cd4$POS <- ifelse(cd4$TSS >= 0, 1, 0)
cd4$afterTSS <- ifelse(cd4$TSS >= 0, cd4$TSS, 0)
```

## 2.  Fit the various representations of time and decide which fits best, while accounting for correlation by including random intercept and serial correlation in your model.  
```{r}
library(nlme)
library(dplyr)

model_a <- lme(CD4 ~ TSS, random = ~1|ID, data = cd4, 
               correlation = corCAR1(form = ~TSS|ID))
model_b1 <- lme(CD4 ~ TSS + POS, random = ~1|ID, data = cd4, 
               correlation = corCAR1(form = ~TSS|ID))
model_b2 <- lme(CD4 ~ TSS + POS + TSS*POS, random = ~1|ID, data = cd4, 
               correlation = corCAR1(form = ~TSS|ID))
model_c <- lme(CD4 ~ TSS + afterTSS, random = ~1|ID, data = cd4, 
               correlation = corCAR1(form = ~TSS|ID))

summary(model_a)
summary(model_b1)
summary(model_b2)
summary(model_c)
AIC(model_a)
AIC(model_b1)
AIC(model_b2)
AIC(model_c)
```

## 3. With your best fitting model, does the effect of time vary across subject? That is, is there a need for a random slope? 
```{r}
model_3 <- lme(CD4 ~ TSS + POS + TSS*POS, random = ~1 + TSS|ID, data = cd4, 
               correlation = corCAR1(form = ~TSS|ID))
summary(model_3)
```

## 4. Is the serial correlation still needed if you have random slopes? Compare the regression
coefficients for time, and the AIC from the model with random intercepts and slopes to that
with random intercepts, slopes, and serial correlation.  
```{r}
model_4 <- lme(CD4 ~ TSS + POS + TSS*POS, random = ~1 + TSS|ID, data = cd4)
summary(model_4)
AIC(model_3); AIC(model_4)
```

## 5. After restricting to the smokers data, is there evidence that smokers have a different change over time?
```{r}
# with random slope 
model_5 <- lme(CD4 ~ TSS + POS + TSS*POS, random = ~1 + TSS|ID, 
               data = filter(cd4, PACKS >0), 
               correlation = corCAR1(form = ~TSS|ID))
summary(model_5)

# without random slope
model_6 <- lme(CD4 ~ TSS + POS + TSS*POS, random = ~1 |ID, 
               data = filter(cd4, PACKS >0), 
               correlation = corCAR1(form = ~TSS|ID))
summary(model_6)

AIC(model_5); AIC(model_6)
```

